---
title: "Diabetes Risk Predictions"
subtitle: "Data Understanding"
author: "Seif Kungulio"
date: "07/11/2025"
output: 
  #pdf_document: default 
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install if needed
#install.packages(c("tidyverse", "skimr", "DataExplorer", "GGally"))

# Load libraries
library(tidyverse)
library(skimr)
library(DataExplorer)
library(GGally)
```





## **Evaluating Dataset**



## **Data Dictionary**



## **Exploratory Data Analysis**
### **Data Loading**
Load the dataset and name it Diabetes.df
```{r}
Diabetes.df <- read.csv("diabetes_health_indicators_BRFSS2015.csv", header = TRUE)
```
Examine the dimension of the data frame
```{r}
dim(Diabetes.df)
```
Preview the data frame
```{r}
glimpse(Diabetes.df)
```
Display the first six rows of the data frame
```{r}
head(Diabetes.df)
```
Display the last six rows of the data frame
```{r}
tail(Diabetes.df)
```
Display the structure of the data frame
```{r}
str(Diabetes.df)
```

### Basic Summary and Structure
Summarize the dataset
```{r}
summary(Diabetes.df)
```
Skim for extended summary
```{r}
#skim(Diabetes.df)
```

## **Data Cleaning**
### Check for Class Imbalance (target Variable)
Distribution of diabetes classes
```{r}
Diabetes.df %>%
  count(Diabetes_binary) %>%
  mutate(Percent = round(n / sum(n) * 100, 2))
```
Plot the distribution
```{r}
ggplot(Diabetes.df,  aes(factor(Diabetes_binary))) +
  geom_bar(fill = "salmon") +
  labs()
```

### Check for Missing Values and Data Quality
Check for missing values
```{r}
colSums(is.na(Diabetes.df))
```

## **EDA: Visualizing Data**
### Univariate Analysis (Numeric Variables)



## **EDA: Determining Relationship**