---
title: "K-Means Clustering"
author: "Seif Kungulio"
date: "04/22/2025"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install necessary packages if not installed
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("cluster", quietly = TRUE)) {
  install.packages("cluster")
}
if (!requireNamespace("factoextra", quietly = TRUE)) {
  install.packages("factoextra")
}

# Load necessary libraries
library(ggplot2)
library(cluster)
library(factoextra)
```
<hr style="height:5px; border:none; color:#006000; background-color:#006000;" />

# **Credit Cards Customer Segmentantion**
<hr style="height:2px; border:none; color:#333; background-color:#333;" />

You were just hired as a data scientist for one of the largest retailer companies: MarketCo. MarketCo offers its customers store credit cards. MarketCo has collected transactions of these credit cards. You are asked by your boss, Mr. Johnson, to develop a customer segmentation to define marketing strategy. You are given a historical dataset summarizing the usage behavior of about 9,000 active credit card holders during the last 6 months. You are asked to perform cluster analysis of this dataset.

**Data Source:** from Kaggle at https://www.kaggle.com/arjunbhasin2013/ccdata/home

## **Data Dictionary**
This dataset contains the following columns:

| **Variable** | **Description** | **Data Type** | **Rules/Constraints** |
|:-------------|:----------------|:--------------|:----------------------|


<br>

## **Question 1**
**Load the dataset CreditCards.csv into memory.**

Read the dataset into memory
```{r}
CreditCards.df <- read.csv("data/CreditCards.csv")
```

Display the dimensions of the data frame (number of rows and columns)
```{r}
dim(CreditCards.df)
```

Display the column names of the data frame
```{r}
colnames(CreditCards.df)
```

<br>

## **Question 2**
**Perform the k-means cluster analysis**

### **Section A**
**Remove the first column: CUST_ID since it doesn’t provide any info for cluster.**

Remove the first column
```{r}
CreditCards.df <- CreditCards.df[,-1]
```

Display the dimensions of the data frame (number of rows and columns)
```{r}
dim(CreditCards.df)
```

Display the column names of the data frame
```{r}
colnames(CreditCards.df)
```

### **Section B**
**Determine the optimal number of clusters. Justify your answer. It may take longer running time since it uses a large dataset.**

For reproducibility
```{r}
set.seed(123)
```

Scale the data for clustering
```{r}
scaled_data <- scale(CreditCards.df)
```

Determine the optimal number of clusters using "gap_stat" Method
```{r, warning=FALSE, message=FALSE}
fviz_nbclust(scaled_data, kmeans, method = "gap_stat") +
  ggtitle("Optimal Number of Clusters - Gap Statistic Method") +
  theme_test() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10)
  )
```


```{r, warning=FALSE, message=FALSE}
# Determine the optimal number of clusters using the Elbow Method
fviz_nbclust(scaled_data, kmeans, method = "wss") +
  ggtitle("Optimal Number of Clusters - Elbow Method") +
  theme_test() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10)
  )
```


```{r, warning=FALSE, message=FALSE}
# Determine the optimal number of clusters using the Silhouette Method
fviz_nbclust(scaled_data, kmeans, method = "silhouette") +
  ggtitle("Optimal Number of Clusters - Silhouette Method") +
  theme_test() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10)
  )
```

Set observed optimal number
```{r}
optimal_k <- 7
```


### **Section C**
**Perform k-means clustering using the optimal number of clusters.**

Perform k-means clustering
```{r, warning=FALSE, message=FALSE}
kmeans_result <- kmeans(scaled_data, centers = optimal_k, nstart = 25)
```

Display the statistical summary of the kmeans
```{r}
summary(kmeans_result)
```


### **Section D**
**Visualize the clusters in different colors.**

Add the cluster assignment to the original data
```{r}
CreditCards.df$Cluster <- as.factor(kmeans_result$cluster)
```

Visualize the clusters
```{r, warning=FALSE, message=FALSE}
fviz_cluster(
  kmeans_result,
  data = scaled_data,
  geom = "point",
  ellipse.type = "norm",
  ggtheme = theme_test(),
  main = "Cluster Visualization"
)
```



