---
title: "Predictive Modeling"
author: "Seif Kungulio"
date: "05/04/2025"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install necessary packages if not installed
if (!requireNamespace("MASS", quietly = TRUE)) {
  install.packages("MASS")
}
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
if (!requireNamespace("caret", quietly = TRUE)) {
  install.packages("caret")
}
if (!requireNamespace("randomForest", quietly = TRUE)) {
  install.packages("randomForest")
}
if (!requireNamespace("e1071", quietly = TRUE)) {
  install.packages("e1071")
}
if (!requireNamespace("cluster", quietly = TRUE)) {
  install.packages("cluster")
}
if (!requireNamespace("factoextra", quietly = TRUE)) {
  install.packages("factoextra")
}
if (!requireNamespace("neuralnet", quietly = TRUE)) {
  install.packages("neuralnet")
}

# Load libraries
library(MASS)
library(tidyverse)
library(caret)
library(tree)
library(randomForest)
library(e1071)
library(cluster)
library(factoextra)
library(neuralnet)
```
<hr style="height:5px; border:none; color:#006000; background-color:#006000;" />

# **Bank of Universe Insurance Health Plans**
<hr style="height:2px; border:none; color:#333; background-color:#333;" />

You are just hired as a Senior Data Scientist in Bank of Universe (BOU). BOU is a private company founded twenty years ago and now has more than 5,000 employees. BOU did all the research, chose the insurance company, and picked plan options for employees twenty years ago. The new CEO, Mr. Buffet, wants to make changes and offer self-funded Health Plans (SHP) starting next year. SHP is cheaper for BOU, since BOU does not have to pay for the separate insurance carrier by taking some risks. BOU has received several years' medical costs in file insurance.csv Download insurance.csvfrom the current insurance carrier.

**Data source:** From Kaggle https://www.kaggle.com/mirichoi0218/insurance/home

It contains the following columns:

+ age: age of primary beneficiary
+ sex: insurance contractor gender, female, male
+ bmi: Body mass index, providing an understanding of body, weights that are relatively high or low relative to height, objective index of body weight (kg / m ^ 2) using the ratio of height to weight, ideally 18.5 to 24.9
+ children: Number of children covered by health insurance / Number of dependents
+ smoker: smoking
+ region: the beneficiary's residential area in the US, northeast, southeast, southwest, northwest.
+ charges: individual medical costs billed by health insurance

<hr><br>

## 1. Data Preparation.
### Section A:
Load the dataset insurance.csv into memory.
```{r}
insurance <- read.csv("data/insurance.csv")

dim(insurance)

head(insurance)
```

### Section B:
In the data frame, transform the variable charges by setting insurance$charges = log(insurance$charges). Do not transform it outside of the data frame.
```{r}
insurance$charges <- log(insurance$charges)
```

Display the first six occurrences
```{r}
head(insurance$charges)
```

### Section C:
Using the data set from 1.b, use the model.matrix() function to create another data set that uses dummy variables in place of categorical variables. Verify that the first column only has ones (1) as values, and then discard the column only after verifying it has only ones as values.
```{r}
insurance_dummy <- model.matrix(charges ~ ., data = insurance)
```

Verify first column has only 1's
```{r}
all(insurance_dummy[, 1] == 1)
```

Display the first ten rows
```{r}
head(insurance_dummy[, 1], 10)
```

Remove the intercept column (first column)
```{r}
insurance_dummy <- insurance_dummy[, -1]
```

### Section D:
Use the sample() function with set.seed equal to 1 to generate row indexes for your training and tests sets, with 2/3 of the row indexes for your training set and 1/3 for your test set. Do not use any method other than the sample() function for splitting your data.

Generate row indexes
```{r}
set.seed(1)
sample_index <- sample(1:nrow(insurance), nrow(insurance) * 2/3)
```

### Section E:
Create a training and test data set from the data set created in 1.b using the training and test row indexes created in 1.d. Unless otherwise stated, only use the training and test data sets created in this step.
```{r}
train_data <- insurance[sample_index, ]
dim(train_data)
test_data <- insurance[-sample_index, ]
dim(test_data)
```

### Section F:
Create a training and test data set from data set created in 1.c using the training and test row indexes created in 1.d
```{r}
train_dummy <- insurance_dummy[sample_index, ]
dim(train_dummy)
test_dummy <- insurance_dummy[-sample_index, ]
dim(test_dummy)
```


<hr><br>

## 2. Build a multiple linear regression model. 
### Section A:
Perform multiple linear regression with charges as the response and the predictors are age, sex, bmi, children, smoker, and region. Print out the results using the summary() function. Use the training data set created in step 1.e to train your model.
```{r}
set.seed(1)
lm_model <- lm(charges ~ age + sex + bmi + children + smoker + region, 
               data = train_data)

summary(lm_model)
```

### Section B:
Is there a relationship between the predictors and the response?
```{r}

```

### Section C:
Does sex have a statistically significant relationship to the response?
```{r}

```

### Section D:
Perform best subset selection using the stepAIC() function from the MASS library, choose best model based on AIC. For the "direction" parameter in the stepAIC() method, set direciton="backward"
```{r}
lm_best <- stepAIC(lm_model, direction = "backward")

summary(lm_best)
```

### Section E:
Compute the test error of the best model in #3d based on AIC using LOOCV using trainControl() and train() from the caret library. Report the MSE by squaring the reported RMSE.
```{r}
ctrl_loocv <- trainControl(method = "LOOCV")

set.seed(1)
lm_loocv <- train(charges ~ age + bmi + children + smoker + region, 
                  data = train_data, 
                  method = "lm", 
                  trControl = ctrl_loocv)

mse_loocv <- lm_loocv$results$RMSE^2

mse_loocv

```

### Section F:
Calculate the test error of the best model in #3d based on AIC using 10-fold Cross-Validation. Use train and trainControl from the caret library. Refer to model selected in #3d based on AIC. Report the MSE.
```{r}
ctrl_cv10 <- trainControl(method = "cv", number = 10)

set.seed(1)
lm_cv10 <- train(charges ~ age + bmi + children + smoker + region, 
                 data = train_data, 
                 method = "lm", 
                 trControl = ctrl_cv10)

mse_cv10 <- lm_cv10$results$RMSE^2

mse_cv10
```

### Section G:
Calculate and report the test MSE using the best model from 2.d and test data set created in step 1.e.
```{r}
pred_lm <- predict(lm_best, newdata = test_data)

mse_test_lm <- mean((test_data$charges - pred_lm)^2)

mse_test_lm
```

### Section H:
Compare the test MSE calculated in step 2.f using 10-fold cross-validation with the test MSE calculated in step 2.g. How similar are they?
```{r}
print(c(LOOCV = mse_loocv, 
        CV10 = mse_cv10, 
        Test = mse_test_lm))
```


<hr><br>

## 3. Build a regression tree model.
### Section A:
Build a regression tree model using function tree(), where charges is the response and the predictors are age, sex, bmi, children, smoker, and region.
```{r}
tree_model <- tree(charges ~ age + sex + bmi + children + smoker + region, 
                   data = train_data)
```

### Section B:
Find the optimal tree by using cross-validation and display the results in a graphic. Report the best size.
```{r}
cv_tree <- cv.tree(tree_model)
plot(cv_tree$size, cv_tree$dev, type = "b", xlab = "Tree Size", ylab = "Deviance")

best_size <- 3
```

### Section C:
Justify  the number you picked for the optimal tree with regard to the principle of variance-bias trade-off.
```{r}

```

### Section D:
Prune the tree using the optinal size found in 3.b.
```{r}
pruned_tree <- prune.tree(tree_model, best = best_size)
```

### Section E:
Plot the best tree model and give labels.
```{r}
plot(pruned_tree)
text(pruned_tree, pretty = 0)
```

### Section F:
Calculate the test MSE for the best model.
```{r}
pred_tree <- predict(pruned_tree, newdata = test_data)

mse_test_tree <- mean((test_data$charges - pred_tree)^2)

mse_test_tree
```


<hr><br>

## 4. Build a random forest model.
### Section A:
Build a random forest model using function randomForest(), where charges is the response and the predictors are age, sex, bmi, children, smoker, and region.
```{r}
set.seed(1)
rf_model <- randomForest(charges ~ age + sex + bmi + children + smoker + region, 
                         data = train_data, importance = TRUE)
```

### Section B:
Compute the test error using the test data set.
```{r}
pred_rf <- predict(rf_model, newdata = test_data)

mse_test_rf <- mean((test_data$charges - pred_rf)^2)

mse_test_rf
```

### Section C:
Extract variable importance measure using the importance() function.
```{r}
importance(rf_model)
```

### Section D:
Plot the variable importance using the function, varImpPlot(). Which are the top 3 important predictors in this model?
```{r}
varImpPlot(rf_model)
```


<hr><br>

## 5. Build a support vector machine model.
### Section A:
The response is charges and the predictors are age, sex, bmi, children, smoker, and region. Please use the svm() function with radial kernel and gamma=5 and cost = 50.
```{r}
set.seed(1)
svm_model <- svm(charges ~ age + sex + bmi + children + smoker + region, 
                 data = train_data, kernel = "radial", gamma = 5, cost = 50)
```

### Section B:
Perform a grid search to find the best model with potential cost: 1, 10, 50, 100 and potential gamma: 1,3 and 5 and potential kernel: "linear","radial" and "sigmoid". And use the training set created in step 1.e.
```{r}
svm_tune <- tune(svm, charges ~ ., data = train_data,
                 ranges = list(kernel = c("linear", "radial", "sigmoid"),
                               cost = c(1, 10, 50, 100),
                               gamma = c(1, 3, 5)))
```

### Section C:
Print out the model results. What are the best model parameters?
```{r}
summary(svm_tune)

best_svm <- svm_tune$best.model

best_svm
```

### Section D:
Forecast charges using the test dataset and the best model found in c).
```{r}
pred_svm <- predict(best_svm, newdata = test_data)
```

### Section E:
Compute the MSE (Mean Squared Error) on the test data.
```{r}
mse_test_svm <- mean((test_data$charges - pred_svm)^2)

mse_test_svm
```


<hr><br>

## 6. Perform the k-means cluster analysis.  
### Section A:
Remove the sex, smoker, and region, since they are not numerical values.
```{r}
insurance_numeric <- insurance[, c("age", "bmi", "children", "charges")]
```

### Section B:
Determine the optimal number of clusters. Justify your answer. It may take longer running time since it uses a large dataset.
```{r}
fviz_nbclust(insurance_numeric, kmeans, method = "gap_stat") +
  ggtitle("Optimal Number of Clusters - Gap Statistic Method") +
  theme_test()
```

### Section C:
Perform k-means clustering using the 3 clusters.
```{r}
kmeans_model <- kmeans(insurance_numeric, centers = 3, nstart = 25)
```

### Section D:
Visualize the clusters in different colors.
```{r}
fviz_cluster(kmeans_model, 
             data = insurance_numeric,
             geom = "point")
```


<hr><br>

## 7. Build a neural networks model.
### Section A:
Remove the sex, smoker, and region, since they are not numerical values.
```{r}
insurance_nn <- insurance[, c("age", "bmi", "children", "charges")]
```

### Section B:
Standardize the inputs using the scale() function.
```{r}
insurance_scaled <- scale(insurance_nn)
```

### Section C:
Convert the standardized inputs to a data frame using the as.data.frame() function.
```{r}
insurance_scaled <- as.data.frame(insurance_scaled)
```

### Section D:
Split the dataset into a training set containing 80% of the original data and the test set containing the remaining 20%.
```{r}
set.seed(1)
index_nn <- sample(1:nrow(insurance_scaled), nrow(insurance_scaled)*0.8)
train_nn <- insurance_scaled[index_nn, ]
test_nn <- insurance_scaled[-index_nn, ]
```

### Section E:
The response is charges and the predictors are age, bmi, and children. Please use 1 hidden layer with 1 neuron.
```{r}
nn_model <- neuralnet(charges ~ age + bmi + children, 
                 data = train_nn, hidden = c(1))
```

### Section F:
Plot the neural networks.
```{r}
plot(nn_model)
```

### Section G:
Forecast the charges in the test dataset.
```{r}
pred_nn <- predict(nn_model, newdata = test_nn)
```

### Section H:
Get the observed charges of the test dataset.
```{r}
obs_nn <- test_nn$charges
```

### Section I:
Compute test error (MSE).
```{r}
mse_test_nn <- mean((obs_nn - pred_nn)^2)

mse_test_nn
```


<hr><br>

## 8. Putting it all together.
### Section A:
For predicting insurance charges, your supervisor asks you to choose the best model among the multiple regression, regression tree, random forest, support vector machine, and neural network models. Compare the test MSEs of the models generated in steps 2.g, 3.f, 4.b, 5.e, and 7.d. Display the names for these types of these models, using these labels: Multiple Linear Regression, Regression Tree, Random Forest, Support Vector Machine, and Neural Network and their corresponding test MSEs in a data.frame. Label the column in your data frame with the labels as Model.Type, and label the column with the test MSEs as Test.MSE and round the data in this column to 4 decimal places. Present the formatted data to your supervisor and recommend which model is best and why.
```{r}
model_compare <- data.frame(
  Model.Type = c("Multiple Linear Regression", "Regression Tree", 
                 "Random Forest", "Support Vector Machine", "Neural Network"),
  Test.MSE = round(c(mse_test_lm, mse_test_tree, mse_test_rf, 
                     mse_test_svm, mse_test_nn), 4)
)
print(model_compare)

# Recommendation based on lowest MSE
best_model <- model_compare[which.min(model_compare$Test.MSE), ]

print(paste("Recommended model is", best_model$Model.Type, 
            "with Test MSE =", best_model$Test.MSE))
```

### Section B:
Another supervisor from the sales department has requested your help to create a predictive model that his sales representatives can use to explain to clients what the potential costs could be for different kinds of customers, and they need an easy and visual way of explaining it. What model would you recommend, and what are the benefits and disadvantages of your recommended model compared to other models?
```{r}

```

### Section C:
The supervisor from the sales department likes your regression tree model. But she says that the salespeople say the numbers in it are way too low and suggests that maybe the numbers on the leaf nodes predicting charges are log transformations of the actual charges. You realize that in step 1.b of this project that you had indeed transformed charges using the log function. And now you realize that you need to reverse the transformation in your final output. The solution you have is to reverse the log transformation of the variables in the regression tree model you created and redisplay the result.

**Follow these steps:**

+ Copy your pruned tree model to a new variable.
```{r}
tree_copy <- pruned_tree
```

+ In your new variable, find the data.frame named "frame" and reverse the log transformation on the data.frame column yval using the exp() function. (If the copy of your pruned tree model is named copy_of_my_pruned_tree, then the data frame is accessed as copy_of_my_pruned_tree$frame, and it works just like a normal data frame.).
```{r}
tree_copy$frame$yval <- exp(tree_copy$frame$yval)
```

+ After you reverse the log transform on the yval column, then replot the tree with labels.
```{r}
plot(tree_copy)
text(tree_copy, pretty = 0)
```

<hr><br>
