---
title: "Final Project"
author: "Seif Kungulio"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### **Question 14**
Summarize Question 13-C using R markdown to generate a reproducible report.  Include the following scripts in your R markdown file:

#### **A. Load the data as specified in Question 1.**

Set the working directory to the correct location for the dataset.
```{r}
setwd("C:/PROJECTS/Maryville/DSCI 502/Week8")
```

Load the data from loan.csv
```{r}
Bikes.df <- read.csv("day.csv")
```

Display the dimensions (rows and columns) of the dataframe
```{r echo=FALSE}
dim(Bikes.df)
```

Display column names of the data frame
```{r echo=FALSE}
colnames(Bikes.df)
```

Display the structure of the data frame
```{r echo=FALSE}
str(Bikes.df)
```

---

<br>

#### **B. Convert the two variables as specified in Question 2.**

Convert numerical season (1,2,3, 4) to characters (springer, summer, fall and winter)
```{r}
Bikes.df$season <- factor(Bikes.df$season,
                          levels = c(1, 2, 3, 4),
                          labels = c("spring", "summer", "fall", "winter"))
```

Display statistical summary of the season variable
```{r summarize-season, echo=FALSE}
summary(Bikes.df$season)
```

Convert numerical weathersit (1,2,3,4) to characters (Good, Mist, Bad, Severe)
```{r convert-weathersit}
Bikes.df$weathersit <- factor(Bikes.df$weathersit,
                              levels = c(1, 2, 3, 4),
                              labels = c("Good", "Mist", "Bad", "Severe"))
```

Display statistical summary of the weathersit variable
```{r summarize-weathersit, echo=FALSE}
summary(Bikes.df$weathersit)
```
---

<br>

#### **C. Convert the categorical variables to factors as specified in Question 3**
Convert categorical variables to factors
```{r}
categorical_vars <- c("season", "holiday", "workingday", "weathersit", "casual")
Bikes.df[categorical_vars] <- lapply(Bikes.df[categorical_vars], as.factor)
```

Display statistical summary of the categorical variables
```{r summarize-categorical-variables, echo=FALSE}
summary(Bikes.df[categorical_vars])
```
---

<br>

#### **D. Build a linear model as specified in Question 13-C. Use R markdown to report the math formula with numerical coefficients for predictors atemp, hum, windspeed, and registered. Skip the coefficients for season, holiday, workingday and  weathersit.**
```{r}

```
---



