---
title: "Final Project"
author: "Seif Kungulio"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Question 14
14 Summarize Question 13-C using R markdown to generate a reproducible report.  Include the following scripts in your R markdown file:

A. Load the data as specified in Question 1.

Set the working directory to the correct location for the dataset.
```{r}
setwd("C:/PROJECTS/Maryville/DSCI 502/Week8")
```

Import necessary libraries
```{r}
#library(dplyr)
#library(ggplot2)
#library(MASS)
#library(knitr)
```
Load the data from loan.csv
```{r}
Bikes.df <- read.csv("day.csv")
```

Display the dimensions (rows and columns) of the dataframe
```{r}
dim(Bikes.df)
```

Display column names of the data frame
```{r}
colnames(Bikes.df)
```

Display the structure of the data frame
```{r}
str(Bikes.df)
```


B. Convert the two variables as specified in Question 2.
```{r}

```

C. Convert the categorical variables to factors as specified in Question 3
```{r}

```

D. Build a linear model as specified in Question 13-C. Use R markdown to report the math formula with numerical coefficients for predictors atemp, hum, windspeed, and registered. Skip the coefficients for season, holiday, workingday and  weathersit.
```{r}

```



