---
title: "Linear Regression Models"
author: "Seif Kungulio"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<hr style="height:5px; border:none; color:#006000; background-color:#006000" />

<br>

## **Dataset Loading**

```{r, echo = FALSE}
# Install necessary packages if not installed
if (!requireNamespace("readxl", quietly = TRUE)) {
  install.packages("readxl")
}

#load the data set into memory
library(readxl)

# Set the working directory
setwd("C:/PROJECTS/Maryville/DSCI-512/Project2")

# Load the Boston dataset
mtcars <- read_excel("data/mtcars.xlsx", sheet = "mtcars")
```

Check the dimension of the data frame
```{r, echo=FALSE}
dim(mtcars)
```

Display data frame column names
```{r, echo=FALSE}
colnames(mtcars)
```

Display data frame structures
```{r, echo=FALSE}
str(mtcars)
```

Display data frame statistical summary
```{r, echo=FALSE}
summary(mtcars)
```
<hr>

## **Question 1**
**Use the lm() function to perform a simple linear regression with the response mpg and the predictor hp.**

Visualize the data pairwise  
```{r fig.width=12, fig.height=8, echo=FALSE}
pairs(mtcars[,c("mpg","hp")])
```

Build the simple linear model  
```{r}
simple_model = lm(mpg ~ hp, data = mtcars)
```

Print out the model results  
```{r, echo=FALSE}
summary(simple_model)
```

<hr>

## **Question 2** 
**Is there a relationship between the target mpg and predator hp?**

**Yes**, the t-test for hp results in a p-value of 2.72e-09, which is significantly low, indicating a strong relationship. Since this is a simple linear regression, the F-test yields the same conclusion:

+ t-value = -7.658
+ Squaring the t-value: (-7.658)^2 = 58.65, which equals the F-value

Thus, the p-values from both the t-test and F-test are identical in this case.


<hr>

## **Question 3**
**How strong is the relationship between the response and predictor?**

The p-value from the t-test is far below the **0.05** threshold, indicating a **strong relationship** between the predictor and the response variable. Additionally, the R^2 value suggests that this variable alone explains **60%** of the variation in the model. The F-test can also serve as a useful indicator.

<hr>

## **Question 4**
**Is the relationship between mpg and hp positive or negative?**

**Negative**, as hp goes up, mpg goes down.

<hr>

## **Question 5**
**What is the predicted mpg associated with a horsepower (hp) of 100? What’s the 95% confidence interval for the predicted mpg?**

Predict mpg for hp = 100 with 95% confidence interval  
```{r}
predicted_mpg <- predict(simple_model, data.frame(hp = 100), 
                         interval = "confidence", level = 0.95)
```

Print the predicted mpg  
```{r, echo=FALSE}
predicted_mpg
```

<hr>

## **Question 6**
**Plot the response and the predictor and add the regression line using abline().**

Plot the response and predictor with the regression line  
```{r fig.width=12, fig.height=8, echo=FALSE}
plot(mtcars$hp, mtcars$mpg, main = "MPG vs Horsepower",
     xlab = "Horsepower (hp)", ylab = "Miles per Gallon (mpg)")
abline(simple_model, lwd = 2, col = "blue")
```

<hr>

## **Question 7**
**Perform a multiple linear regression with mpg as the response and the predictors cyl, disp, hp, wt, vs, and gear. Print out the results using summary() function.**

Perform multiple linear regression  
```{r}
multiple_model <- lm(mpg ~ cyl + disp + hp + wt + vs + gear, data = mtcars)
```

Print the summary of the model  
```{r, echo=FALSE}
summary(multiple_model)
```

<hr>

## **Question 8**
**Is there a relationship between the predictors and the response?**

**Yes**, some predictors show a significant relationship with the response variable, as evidenced by the F-statistic of 30.5 and a p-value of 2.568e-12.

<hr>

## **Question 9**
**Which predictors appears to have a statistically significant relationship to the response?**

The variables **wt** (weight) and **hp** (gross horsepower) have p-values of **0.00012** and **0.02**, respectively, based on their t-tests.

<hr>

## **Question 10**
**Use * symbols to fit linear regression models with interaction effects between hp and wt. Does this interaction appear to be statistically significant?**

Fit a model with interaction effects between hp and wt  
```{r}
interaction_model <- lm(mpg ~ hp * wt, data = mtcars)
```

Print the summary of the interaction model  
```{r, echo=FALSE}
summary(interaction_model)
```

**Yes**, the interaction has a p-value of 0.000362 < alpha = 0.05

<hr>
