---
title: "Diabetes Risks Prediction Dashboard"
output: 
  flexdashboard::flex_dashboard:
    theme:
      bg: "#131A22"
      #bg: "#101010"
      fg: "#FDF7F7" 
      primary: "#ED79F9"
      base_font:
        google: Prompt
      code_font:
        google: JetBrains Mono
    orientation: rows
    vertical_layout: fill
runtime: shiny
---


```{r setup, include=FALSE}
# ---- Packages ----
suppressPackageStartupMessages({
  library(flexdashboard)   # layout + valueBox
  library(shiny)           # interactivity
  library(tidyverse)       # dplyr, tidyr, ggplot2, readr
  library(scales)          # percent labels
  library(treemapify)      # treemap
  library(plotly)          # hover tooltips for ggplot
  library(knitr)           # kable
  library(kableExtra)      # kable styling
  library(DT)              # data table
})

options(scipen = 999)
theme_set(theme_minimal(base_size = 13))

# ---- Helper: read data ----
# The app reads from a hard-coded path (no upload UI).
default_path <- "diabetes_data.csv"   # <- replace with your local path if desired

read_diabetes <- function(path) {
  readr::read_csv(path, show_col_types = FALSE)
}

# ---- Reactive data source ----
raw_data <- reactive({
  req(file.exists(default_path))
  read_diabetes(default_path)
})

# ---- Clean/select variables + derive fields ----
base_df <- reactive({
  df <- raw_data()
  
  # Ensure expected columns exist
  expected <- c("Diabetes_binary","BMI","BMI_scaled","MentHlth_binned","PhysHlth_binned",
                "Age_group","Chronic_Risk_Load","Healthcare_Barrier_Index","GenHlth",
                "Education","Income","Smoker","PhysActivity","DiffWalk",
                "HvyAlcoholConsump","Sex")
  missing <- setdiff(expected, names(df))
  validate(
    need(length(missing) == 0, paste0("Missing columns: ", paste(missing, collapse = ", ")))
  )

  df <- df %>%
    dplyr::select(dplyr::all_of(expected)) %>%
    # make strings/factors consistent
    mutate(
      across(c(Age_group, GenHlth, Education, Income, Sex), as.factor),
      across(c(Smoker, PhysActivity, DiffWalk, HvyAlcoholConsump, Diabetes_binary), ~as.integer(.x))
    )
  
  # Recover approximate BMI from scaled BMI if present; else use BMI directly
  if ("BMI_scaled" %in% names(df) && all(is.finite(df$BMI)) && sd(df$BMI, na.rm = TRUE) > 0) {
    mean_bmi <- mean(df$BMI, na.rm = TRUE)
    sd_bmi   <- sd(df$BMI, na.rm = TRUE)
    df$approx_BMI <- (df$BMI_scaled * sd_bmi) + mean_bmi
  } else {
    df$approx_BMI <- df$BMI
  }
  
  df
})

# ---- Filtered dataset according to inputs ----
filtered_df <- reactive({
  df <- base_df()
  
  # Apply sidebar filters (reduced set)
  if (!is.null(input$age)) df <- df %>% filter(is.na(Age_group) | Age_group %in% input$age)
  if (!is.null(input$sex)) df <- df %>% filter(is.na(Sex) | Sex %in% input$sex)
  if (!is.null(input$edu)) df <- df %>% filter(is.na(Education) | Education %in% input$edu)
  if (!is.null(input$inc)) df <- df %>% filter(is.na(Income) | Income %in% input$inc)
  if (!is.null(input$gh))  df <- df %>% filter(is.na(GenHlth) | GenHlth %in% input$gh)
  if (!is.null(input$smk)) df <- df %>% filter(is.na(Smoker) | Smoker %in% input$smk)
  
  # numeric/range filters
  if (!is.null(input$crl))    df <- df %>% filter(is.na(Chronic_Risk_Load) | dplyr::between(Chronic_Risk_Load, input$crl[1], input$crl[2]))
  if (!is.null(input$hbi))    df <- df %>% filter(is.na(Healthcare_Barrier_Index) | dplyr::between(Healthcare_Barrier_Index, input$hbi[1], input$hbi[2]))
  if (!is.null(input$bmirng)) df <- df %>% filter(is.na(approx_BMI) | dplyr::between(approx_BMI, input$bmirng[1], input$bmirng[2]))
  
  # optional: exclude known diabetes cases from pool (so HighRisk compares among non-diabetics)
  if (isTRUE(input$exclude_known)) {
    df <- df %>% filter(Diabetes_binary == 0)
  }
  
  # High-risk flag based on adjustable BMI cutoff
  df <- df %>% mutate(HighRisk = if_else(approx_BMI > input$bmi_cutoff & Diabetes_binary == 0, 1L, 0L))
  df
})

# ---- Initialize dynamic choices when data loads ----
observeEvent(base_df(), {
  df <- base_df()
  updateSelectizeInput(inputId = "age", choices = sort(unique(na.omit(df$Age_group))), server = TRUE)
  updateSelectizeInput(inputId = "sex", choices = sort(unique(na.omit(df$Sex))), server = TRUE)
  updateSelectizeInput(inputId = "edu", choices = sort(unique(na.omit(df$Education))), server = TRUE)
  updateSelectizeInput(inputId = "inc", choices = sort(unique(na.omit(df$Income))), server = TRUE)
  updateSelectizeInput(inputId = "gh",  choices = sort(unique(na.omit(df$GenHlth))), server = TRUE)
  updateSelectizeInput(inputId = "smk", choices = sort(unique(na.omit(df$Smoker))), server = TRUE)
  
  # sensible numeric ranges
  updateSliderInput(inputId = "crl",
                    min = floor(min(df$Chronic_Risk_Load, na.rm = TRUE)),
                    max = ceiling(max(df$Chronic_Risk_Load, na.rm = TRUE)),
                    value = c(floor(min(df$Chronic_Risk_Load, na.rm = TRUE)),
                              ceiling(max(df$Chronic_Risk_Load, na.rm = TRUE))))
  updateSliderInput(inputId = "hbi",
                    min = floor(min(df$Healthcare_Barrier_Index, na.rm = TRUE)),
                    max = ceiling(max(df$Healthcare_Barrier_Index, na.rm = TRUE)),
                    value = c(floor(min(df$Healthcare_Barrier_Index, na.rm = TRUE)),
                              ceiling(max(df$Healthcare_Barrier_Index, na.rm = TRUE))))
  updateSliderInput(inputId = "bmirng",
                    min = floor(min(df$approx_BMI, na.rm = TRUE)),
                    max = ceiling(max(df$approx_BMI, na.rm = TRUE)),
                    value = c(18, 45))
})

# KPI data
kpi <- reactive({
  df <- filtered_df()
  tibble(
    pct_high_risk     = mean(df$HighRisk == 1, na.rm = TRUE) * 100,
    total_high_risk   = sum(df$HighRisk == 1, na.rm = TRUE),
    avg_BMI_high_risk = mean(df$approx_BMI[df$HighRisk == 1], na.rm = TRUE),
    avg_chronic_load  = mean(df$Chronic_Risk_Load[df$HighRisk == 1], na.rm = TRUE)
  )
})
```

Sidebar {.sidebar}
--------------------------------------------------------------------------------

#### Filters
```{r}
checkboxInput("exclude_known", "Exclude known diabetes cases", value = TRUE)
numericInput("bmi_cutoff", "BMI threshold for High-Risk (>=):", value = 30, min = 18, max = 60, step = 0.5)

selectizeInput("age", "Age group", choices = NULL, multiple = TRUE, options = list(placeholder = "All"))
selectizeInput("sex", "Sex",       choices = NULL, multiple = TRUE, options = list(placeholder = "All"))
selectizeInput("edu", "Education", choices = NULL, multiple = TRUE, options = list(placeholder = "All"))
selectizeInput("inc", "Income",    choices = NULL, multiple = TRUE, options = list(placeholder = "All"))

actionButton("reset", "Reset filters")

observeEvent(input$reset, {
  updateSelectizeInput(inputId = "age", selected = character(0))
  updateSelectizeInput(inputId = "sex", selected = character(0))
  updateSelectizeInput(inputId = "edu", selected = character(0))
  updateSelectizeInput(inputId = "inc", selected = character(0))
})
```

Column {data-height=130, .tabset}
--------------------------------------------------------------------------------

### KPI Overview

```{r}

uiOutput("kpi_table")

# HTML table of KPIs
output$kpi_table <- renderUI({
  kb <- kpi()
  tbl <- kb %>%
    mutate(across(everything(), ~ifelse(is.finite(.), ., NA))) %>%
    kable(format = "html", digits = 2, caption = "KPI Summary", align = "c") %>%
    kable_styling(full_width = TRUE, 
                  bootstrap_options = c("striped", "hover"),
                  font_size = 20)
  HTML(as.character(tbl))
})
```

### Problem Statement

Public health agencies in Texas need a fast way to spot adults **at or above 30%** risk of developing diabetes within 5 years, using an accessible proxy. This dashboard uses **BMI ≥ adjustable threshold (default 30)** among adults who **do not currently have diabetes** to flag **High-Risk** individuals, then lets you slice by **age, sex, and socio-economic indicators** to target **preventive outreach**.


Column {.tabset}
-----------------------------------------------------------------------

### % High Risk

```{r}
plotlyOutput("age_sex_plot")

output$age_sex_plot <- renderPlotly({
  df <- filtered_df(); req(nrow(df) > 0)
  p1 <- ggplot(df, aes(x = Age_group, fill = Sex)) +
    geom_bar(position = "fill", na.rm = TRUE) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "% Distribution by Age and Sex",
      x = "Age Group", y = "Percentage within group", fill = "Sex"
    ) +
    theme_test(base_size = 13)
  ggplotly(p1)
})
```

### High-Risk Prevalence

```{r}
plotOutput("treemap_plot", height = "600px")

output$treemap_plot <- renderPlot({
  df <- filtered_df(); req(nrow(df) > 0)
  df_treemap <- df %>%
    group_by(Education, Income) %>%
    summarise(
      count = dplyr::n(),
      pct_high_risk = mean(HighRisk == 1, na.rm = TRUE) * 100,
      .groups = "drop"
    )

  ggplot(df_treemap, aes(area = count, fill = pct_high_risk,
                         label = paste(Education, "\n", Income,
                                       "\n", sprintf("%.1f%%",
                                                     pct_high_risk)))) +
    geom_treemap() +
    geom_treemap_text(colour = "white", place = "centre", grow = TRUE, reflow = TRUE) +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(title = "High-Risk % by Education & Income", fill = "% High Risk") +
    theme_test(base_size = 13)
})
```

### General Health Ratings

```{r}
plotlyOutput("genhlth_plot")

output$genhlth_plot <- renderPlotly({
  df <- filtered_df(); req(nrow(df) > 0)
  df <- df %>%
    mutate(GenHlth = factor(GenHlth,
                            labels = c("Excellent", "Very Good", 
                                       "Good", "Fair", "Poor")))
  
  p4 <- ggplot(df, aes(x = GenHlth, fill = factor(HighRisk,
                                                  levels = c(0,1),
                                                  labels = c("No", "Yes")))) +
    geom_bar(position = "fill", na.rm = TRUE) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "General Health Ratings vs Risk Status",
      x = "General Health", y = "Percentage", fill = "High Risk"
    ) +
    theme_test(base_size = 13)
  ggplotly(p4)
})
```
